### 2017-08-30 category-theory-basics.tex
Add texts

@@

\begin{definition}
  A \textbf{Lambek category} (or a \textbf{biclosed monoidal category}) is a monoidal category
  $(\cat{M},\tri,I,\alpha,\lambda,\rho)$ equipped with two bifunctors
  $\rightharpoonup:\cat{M}^{op}\times\cat{M}\rightarrow\cat{M}$ and
  $\leftharpoonup:\cat{M}\times\cat{M}^{op}\rightarrow\cat{M}$ that are both right adjoint to
  the tensor product. That is, the following natural bijections hold:
  \begin{center}
  \begin{math}
  \begin{array}{lll}
    \Hom{L}{X\tri A}{B}\cong\Hom{L}{X}{A\lto B} & \quad\quad\quad\quad & 
    \Hom{L}{A\tri X}{B}\cong\Hom{L}{X}{B\rto A}
  \end{array}
  \end{math}
  \end{center}
\end{definition}

@@

\begin{lemma}
  \label{lemma:internal-homs-collapse}
  Let $A$ and $B$ be two objects in a Lambek category with the exchange natural transformation.
  Then $(A \lto B) \cong (B \rto A)$.
\end{lemma}
\begin{proof}
  First, notice that for any object $C$ we have
  \begin{center}
  \begin{math}
  \small
  \begin{array}{lllll}
    Hom[C,A\lto B]
    & \cong & Hom[C\otimes A,B] & \cat{L}\text{ is a Lambek category}\\
    & \cong & Hom[A\otimes C,B] & \text{By the exchange }\e{C,A}\\
    & \cong & Hom[C,B\rto A]    & \cat{L}\text{ is a Lambek category}
  \end{array}
  \end{math}
  \end{center}  
  Thus, $A\lto B\cong B\rto A$ by the Yoneda lemma.
\end{proof}
\begin{corollary}
  \label{corollary:LC-with-ex-mc}
  A Lambek category with exchange is symmetric monoidal closed.
\end{corollary}

\begin{definition}
  Let $(\cat{M},\tri,I,\alpha,\lambda,\rho)$ and
  $(\cat{M'},\tri',I',\alpha',\lambda',\rho')$ be monoidal categories. A \textbf{monoidal
  functor} $(F,\m{})$ from $\cat{M}$ to $\cat{M'}$ is a functor $F:\cat{M}\rightarrow\cat{M'}$
  together with a morphism $\m{I}:I'\rightarrow F(I)$ and a natural transformation
  $\m{A,B}:FA'\tri FB'\rightarrow F(A\tri B)$, such that the following diagrams commute
  for any objects $A$, $B$, and $C$ in $\cat{M}$:
  \begin{mathpar}
  \bfig
    \hSquares/->`->`->``->`->`->/<400>[
      (FA\tri'FB)\tri'FC`FA\tri'(FB\tri'FC)`FA\tri'F(B\tri C)`
      F(A\tri B)\tri'FC`F((A\tri B)\tri C)`F(A\tri(B\tri C));
      \alpha'_{FA,FB,FC}`id_{FA}\tri'\m{A,B}`\m{A,B}\tri'id_{FC}``
      \m{A,B\tri C}`\m{A\tri B,C}`F\alpha_{A,B,C}]
  \efig
  \and
  \bfig
    \square/->`->`<-`->/<600,400>[
      I'\tri'FA`FA`FI\tri'FA`F(I\tri A);
      \lambda'_{FA}`\m{I}\tri id_{FA}`F\lambda_A`\m{I,A}]
  \efig
  \and
  \bfig
    \square/->`->`<-`->/<600,400>[
      FA\tri'I'`FA`FA\tri'FI`F(A\tri I);
      \rho'_{FA}`\id_{FA}\tri\m{I}`F\rho_A`\m{A,I}]
  \efig
  \end{mathpar}
\end{definition}

@@

\begin{definition}
  Let $(\cat{M},\tri,I,\alpha,\lambda,\rho)$ and
  $(\cat{M'},\tri',I',\alpha',\lambda',\rho')$ be monoidal categories,
  $F:\cat{M}\rightarrow\cat{M'}$ and $G:\cat{M}'\rightarrow\cat{M}$ be functors. The adjunction
  $F:\cat{M}\dashv\cat{M'}:G$ is a \textbf{monoidal adjunction} if $F$ and $G$ are monoidal
  functors, and the unit $\eta$ and the counit $\varepsilon$ are monoidal natural
  transformations.
\end{definition}

\begin{definition}
  Let $\cat{C}$ be a category. A \textbf{monad} on $\cat{C}$ consists of an endofunctor
  $T:\cat{C}\rightarrow\cat{C}$ together with two natural transformations
  $\eta:id_\cat{C}\rightarrow T$ and $\mu:T^2\rightarrow T$, where $id_\cat{C}$ is the identity
  functor on $\cat{C}$, such that the following diagrams commute:
  \begin{mathpar}
  \bfig
    \square<400,400>[T^3`T^2`T^2`T;T\mu`\mu_T`\mu`\mu]
  \efig
  \and
  \bfig
    \square<400,400>[T`T^2`T^2`T;\eta_T`T\eta`\mu`\mu]
    \morphism(0,400)/=/<400,-400>[T`T;]
  \efig
  \end{mathpar}
\end{definition}

>>>>

This section contains the basic definitions in category theory that we will be using in our
adjoint model. Our model is based on special kinds of monoidal categories: Lambek categories and
symmetric monoidal closed categories and Lambek categories, as defined in
Definitions~\ref{def:lambek-cat} and~\ref{def:smcc}.

@@

\begin{definition}
\label{def:lambek-cat}
  A \textbf{Lambek category} (or a \textbf{biclosed monoidal category}) is a monoidal category
  $(\cat{M},\tri,I,\alpha,\lambda,\rho)$ equipped with two bifunctors
  $\rightharpoonup:\cat{M}^{op}\times\cat{M}\rightarrow\cat{M}$ and
  $\leftharpoonup:\cat{M}\times\cat{M}^{op}\rightarrow\cat{M}$ that are both right adjoint to
  the tensor product. That is, the following natural bijections hold:
  \begin{center}
  \begin{math}
  \begin{array}{lll}
    \Hom{L}{X\tri A}{B}\cong\Hom{L}{X}{A\lto B} & \quad\quad\quad\quad & 
    \Hom{L}{A\tri X}{B}\cong\Hom{L}{X}{B\rto A}
  \end{array}
  \end{math}
  \end{center}
\end{definition}

@@

The relation between SMMCs and Lambek categories are demonstrated in
Lemma~\ref{lemma:internal-homs-collapse} and Corollary~\ref{corollary:LC-with-ex-mc}.

\begin{lemma}
  \label{lemma:internal-homs-collapse}
  Let $A$ and $B$ be two objects in a Lambek category with the exchange natural transformation.
  Then $(A \lto B) \cong (B \rto A)$.
\end{lemma}
\begin{proof}
  First, notice that for any object $C$ we have
  \begin{center}
  \begin{math}
  \small
  \begin{array}{lllll}
    Hom[C,A\lto B]
    & \cong & Hom[C\otimes A,B] & \cat{L}\text{ is a Lambek category}\\
    & \cong & Hom[A\otimes C,B] & \text{By the exchange }\e{C,A}\\
    & \cong & Hom[C,B\rto A]    & \cat{L}\text{ is a Lambek category}
  \end{array}
  \end{math}
  \end{center}  
  Thus, $A\lto B\cong B\rto A$ by the Yoneda lemma.
\end{proof}

\begin{corollary}
  \label{corollary:LC-with-ex-mc}
  A Lambek category with exchange is symmetric monoidal closed.
\end{corollary}

The essential component in our non-commutative adjoint model is a monoidal adjunction, defined
in Definitions~\ref{def:monoidal-functor}-\ref{def:monoidal-adjunction}.

\begin{definition}
\label{def:monoidal-functor}
  Let $(\cat{M},\tri,I,\alpha,\lambda,\rho)$ and
  $(\cat{M'},\tri',I',\alpha',\lambda',\rho')$ be monoidal categories. A \textbf{monoidal
  functor} $(F,\m{})$ from $\cat{M}$ to $\cat{M'}$ is a functor $F:\cat{M}\rightarrow\cat{M'}$
  together with a morphism $\m{I}:I'\rightarrow F(I)$ and a natural transformation
  $\m{A,B}:FA'\tri FB'\rightarrow F(A\tri B)$, such that the following diagrams commute
  for any objects $A$, $B$, and $C$ in $\cat{M}$:
  \begin{mathpar}
  \bfig
    \hSquares/->`->`->``->`->`->/<400>[
      (FA\tri'FB)\tri'FC`FA\tri'(FB\tri'FC)`FA\tri'F(B\tri C)`
      F(A\tri B)\tri'FC`F((A\tri B)\tri C)`F(A\tri(B\tri C));
      \alpha'_{FA,FB,FC}`id_{FA}\tri'\m{A,B}`\m{A,B}\tri'id_{FC}``
      \m{A,B\tri C}`\m{A\tri B,C}`F\alpha_{A,B,C}]
  \efig
  \and
  \bfig
    \square/->`->`<-`->/<600,400>[
      I'\tri'FA`FA`FI\tri'FA`F(I\tri A);
      \lambda'_{FA}`\m{I}\tri id_{FA}`F\lambda_A`\m{I,A}]
  \efig
  \and
  \bfig
    \square/->`->`<-`->/<600,400>[
      FA\tri'I'`FA`FA\tri'FI`F(A\tri I);
      \rho'_{FA}`\id_{FA}\tri\m{I}`F\rho_A`\m{A,I}]
  \efig
  \end{mathpar}
\end{definition}

@@

\begin{definition}
\label{def:monoidal-adjunction}
  Let $(\cat{M},\tri,I,\alpha,\lambda,\rho)$ and
  $(\cat{M'},\tri',I',\alpha',\lambda',\rho')$ be monoidal categories,
  $F:\cat{M}\rightarrow\cat{M'}$ and $G:\cat{M}'\rightarrow\cat{M}$ be functors. The adjunction
  $F:\cat{M}\dashv\cat{M'}:G$ is a \textbf{monoidal adjunction} if $F$ and $G$ are monoidal
  functors, and the unit $\eta$ and the counit $\varepsilon$ are monoidal natural
  transformations.
\end{definition}

In Moggi's monad model \cite{}, the monad is required to be strong, as defined in
Definitions~\ref{def:monad} and~\ref{def:strong-monad}.

\begin{definition}
\label{def:monad}
  Let $\cat{C}$ be a category. A \textbf{monad} on $\cat{C}$ consists of an endofunctor
  $T:\cat{C}\rightarrow\cat{C}$ together with two natural transformations
  $\eta:id_\cat{C}\rightarrow T$ and $\mu:T^2\rightarrow T$, where $id_\cat{C}$ is the identity
  functor on $\cat{C}$, such that the following diagrams commute:
  \begin{mathpar}
  \bfig
    \square<400,400>[T^3`T^2`T^2`T;T\mu`\mu_T`\mu`\mu]
  \efig
  \and
  \bfig
    \square<400,400>[T`T^2`T^2`T;\eta_T`T\eta`\mu`\mu]
    \morphism(0,400)/=/<400,-400>[T`T;]
  \efig
  \end{mathpar}
\end{definition}
